---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
---

Brief Overview of Data
The data for this assignment can be downloaded from the course web site:

Dataset: Activity monitoring data
The variables included in this dataset are:

steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)

date: The date on which the measurement was taken in YYYY-MM-DD format

interval: Identifier for the 5-minute interval in which measurement was taken
-----------------


## Loading and preprocessing the data

```{r, echo=FALSE}
#Set the Working Directory e.g.
#setwd("~/GitHub/RepData_PeerAssessment1")

#Install and load below packages

#install.packages("ggplot2")
#install.packages("dplyr")
#install.packages("chron") #Used for is.weekend() function

library(ggplot2)
library(dplyr)
library(chron)   #Used for is.weekend() function

```

Unzip data to Obtain csv file

```{r,echo = FALSE}
if(!file.exists('activity.csv')){
  unzip('activity.zip')
}
```
Reading the data into activuity data frame and show some summary statistics

```{r,echo = FALSE}
a <- read.csv("activity.csv",header=T)
str(a)
summary(a)
```

## Process/transform the data (if necessary) into a format suitable for your analysis
```{r,echo = FALSE}
head(a)
```

## What is mean total number of steps taken per day?
1. Calculate the number of steps taken per day (Ignore the missin values)
```{r,echo = FALSE}
#Aggregating(summation) of steps over date
aggsteps<- aggregate(steps ~ date, a, FUN=sum)

#Aggregated Data (all steps added for a particular date)
head(aggsteps)
```
2. Histogram of the total number of steps taken each day
```{r, echo=FALSE}
#Ploting histogram using hist() from Base Plotting
hist(aggsteps$steps, 
     col="red", 
     xlab = "Frequency", 
     ylab = "Steps",
     main = "Total Number Of Steps Taken Each day")
```

 3.Calculate and report the mean and median total number of steps taken per day
 
```{r, echo=FALSE}
amean <- mean(aggsteps$steps)
amedian <- median(aggsteps$steps)

#Mean total number of steps taken per day
amean

#Median total number of steps taken per day
amedian
 
```
 

## What is the average daily activity pattern?
1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r, echo = FALSE}
#Aggregating(summation) of steps over time interval (of 5 min)
agginterval <- aggregate(steps ~ interval, a, FUN=sum)

#Plotting line graph using plot() from Base Plotting for Total Steps vs 5-Minute Interval
plot(agginterval$interval, agginterval$steps, 
     type = "l", lwd = 2,
     xlab = "Interval", 
     ylab = "Total Steps",
     main = "Total Steps vs. 5-Minute Interval")
```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r, echo = FALSE}
#Fetching the 5 min interval which has maximum number of steps
filter(agginterval, steps==max(steps))
```



## Imputing missing values
1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r, echo = FALSE}

#In the Output of the below query TRUE represents the total number of NA values
table(is.na(a))

```

2.Devise a strategy for filling in all of the missing values in the dataset. Create a new dataset that is equal to the original dataset but with the missing data filled in. I will use the mean for that 5-minute interval.

```{r, echo = FALSE}
#In the original data set aggregating (mean) steps over 5-minute interval
meaninterval<- aggregate(steps ~ interval, a, FUN=mean)

#Merging the mean of total steps for a date with the original data set
anew <- merge(x=a, y=meaninterval, by="interval")

#Replacing the NA values with the mean for that 5-minute interval
anew$steps <- ifelse(is.na(anew$steps.x), anew$steps.y, anew$steps.x)

#Merged dataset which will be subsetted in the next step by removing not required columns
head(anew)
```

3.Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r,echo=FALSE}
#Fetching only the required columns (steps, date, interval) and storing in the new data set.
anew <- select(anew, steps, date, interval)

#New dataset with NA imputed by mean for that 5-minute interval
head(anew)
```
# Histogram of the total number of steps taken each day
```{r,echo=FALSE}
#Aggregating(summation) of steps over date
aggsteps_new<- aggregate(steps ~ date, anew, FUN=sum)

#Plotting
#Setting up the pannel for one row and two columns
par(mfrow=c(1,2))

#Histogram after imputing NA values with mean of 5-min interval
hist(aggsteps_new$steps, 
     col="green",
     xlab = "Steps", 
     ylab = "Frequency",
     ylim = c(0,35),
     main = "Total Number Of Steps Taken Each day \n(After imputing NA values with \n mean of 5-min interval)",
     cex.main = 0.7)

#Histogram with the orginal dataset
hist(aggsteps$steps, 
     col="red", 
     xlab = "Steps", 
     ylab = "Frequency",
     ylim = c(0,35),
     main = "Total Number Of Steps Taken Each day \n(Orginal Dataset)",
     cex.main = 0.7)

```

Mean and median number of steps taken each day

```{r,echo=FALSE}
par(mfrow=c(1,1)) #Resetting the panel

amean_new <- mean(aggsteps_new$steps)
amedian_new <- median(aggsteps_new$steps)

#Comparing Means
paste("New Mean      :", round(amean_new,2), "," ,  
      " Original Mean :", round(amean,2),"," , 
      " Difference :",round(amean_new,2) -  round(amean,2))

#Comparing Medians
paste("New Median    :", amedian_new, ",", 
      " Original Median :", amedian,"," , 
      " Difference :",round(amedian_new-amedian,2))

```



## Are there differences in activity patterns between weekdays and weekends?

For this part the weekdays() function may be of some help here. Use the dataset with the filled-in missing values for this part.

1.Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r, echo = FALSE}
#Below code will use "chron" package, please install if not alrady installed
#install.packages("chron")
#library(chron)

#is.weekend() function considers Saturday and Sunday as weekends
#In the output of below query FALSE means weekday, TRUE means weekend
table(is.weekend(anew$date))

#Adding new factor variable "dayofweek" indicating whether a given date is a weekday or weekend day
anew$dayofweek <- ifelse(is.weekend(anew$date), "weekend", "weekday")

#Number of Weekdays and Weekends
table(anew$dayofweek)

#New Data after adding factor variable for weekday or weekend
head(anew)
```


2.Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis)

```{r,echo=TRUE}
#Aggregating(mean) steps over interval and day of week
meaninterval_new<- aggregate(steps ~ interval + dayofweek, anew, FUN=mean)

#Aggregated Data
head(meaninterval_new)

#Time Series plot using ggplot
ggplot(meaninterval_new, aes(x=interval, y=steps)) + 
  geom_line(color="blue", size=1) + 
  facet_wrap(~dayofweek, nrow=2) +
  labs(x="\nInterval", y="\nNumber of steps")
```
